Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-04-05 paper/AlignedIsotropicGW.tex
Update AlignedIsotropicGW.tex
Very small changes; added Kulkarni, Hut, and McMillan 1993 to references regarding black hole dynamics in dense stellar environments

The spin parameter with the largest effect on waveforms, and a
correspondingly tight constraint from the data, is a mass-weighted,
aligned, ``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\left| \vec{L} \right|} = \frac{1}{M} \left( m_1 \chi_1 + m_2 \chi_2 \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $0 \leq \chi_{1,2} \leq 1$ are the corresponding dimensionless
projections of the individual \ac{BH} spins
\citep{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  There is essentially no posterior support for
$\chi \gtrsim 0.5$ as observed in the majority of the
electromagnetically-detected stellar-mass black hole population (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

@@

Small values of $\chieff$ as exhibited in these systems can result
from either intrinsically small spins or larger spins whose direction
is mis-aligned with the orbital plane of the binary (i.e.\ spin
vectors with small $z$-components).  Mis-alignment is capable of
producing \emph{negative} values of $\chieff$, however, while aligned
spins will always have $\chieff \geq 0$.  This difference provides
strong discriminating power between the two angular distributions,
even without good information about the magnitude distribution; to the
extent that data favour negative $\chieff$ they weigh heavily against
aligned models.  To quantify the degree of support for these two
alternate explanations of small $\chieff$ values in the merging
\ac{BBH} population, we compared the Bayesian evidence for various
simple models of the spin population using the \ac{GW} data set.

@@

\begin{figure}
  \plotone{../plots/chi-eff-distributions}
  \caption{\label{fig:chieff-distribution-models} The models for the
    distribution of $\chieff$ considered in this paper.  In all models
    we assume that the binary mass ratio $q \equiv m_1/m_2 = 1$ and
    that the distribution of spin vectors is the same for each
    component.  The ``flat'' (blue lines), ``high,'' (green lines),
    and ``low'' (red lines) magnitude distributions are defined in
    Eq.\ \eqref{eq:magnitude-dists}.  Solid lines give the $\chieff$
    distribution under the assumption that the orientations of the
    spins are isotropic; dashed lines give the distribution under the
    assumption that both objects' spins are aligned with the orbital
    angular momentum.  The isotropic distributions are readily
    distinguished from the aligned by the production of negative
    $\chieff$ values, while the distinction between the three models
    for the spin magnitude distribution is less sharp.}
\end{figure}

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections with similar observational uncertainties
drawn from each of our six zero-parameter models for the spin
distribution in addition to the three existing detections from O1, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
correct angular distribution to be distinguished with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we will remain much less certain
about the \emph{magnitude} distribution until we have a larger number
of observations.  In Figure \ref{fig:O2-predictions}, the odds ratio
between different magnitude distributions with the same angular
distribution is much closer to one than the odds ratio between angular
distributions.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments
\citep{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016}
are expected to be isotropic given the absence of a preferred
direction \citep[e.g.,][]{2016ApJ...832L...2R} and the persistence of
an isotropic distribution through post-Newtonian evolution
\citep{2004PhRvD..70l4020S,2007ApJ...661L.147B}.

The spin directions in isolated binaries, whether evolving via the
classical channel through a common-envelope phase
\citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017}
or through chemically homogeneous evolution
\citep{MandeldeMink:2016,Marchant:2016} are generally expected to be
preferentially aligned.  Despite observed spin-orbit misalignments in
both massive stellar binaries
\citep{Albrecht:2009,2014ApJ...785...83A} and BH XRBs
\citep{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014},
mass accretion and tidal interactions will tend to realign the binary.
On the other hand, a supernova natal kick (if any) can change the
orbital plane and misalign the binary
\citep{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can
also tilt the spin angle, as in the double pulsar J0737-3039
\citep{2011ApJ...742...81F}; and a variety of uncertain processes,
such as wind-driven mass loss and post-collapse fallback, can couple
the spin magnitude and direction distributions, contrary to our
simplified assumptions.

>>>>

The spin parameter with the largest effect on waveforms, and a
correspondingly tight constraint from the data, is a mass-weighted,
aligned, ``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\left| \vec{L} \right|} = \frac{1}{M} \left( m_1 \chi_1 + m_2 \chi_2 \right),
\end{equation}
where the $\hat{z}$ axis is aligned with the orbital angular momentum
of the binary, $m_{1,2}$ are the gravitational masses of the more-massive (1) and
less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
and $0 \leq \chi_{1,2} \leq 1$ are the corresponding dimensionless
projections of the individual \ac{BH} spins
\citep{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from \citet{O1-BBH}.  Because samples
drawn from the posterior on $\chieff$ are not publicly released at
this time, we have approximated the posterior as a Gaussian
distribution with the same mean and 90\% credible interval as quoted
in \citet{O1-BBH}.  There is essentially no posterior support for
$\chi \gtrsim 0.5$ as inferred for the majority of the
electromagnetically-detected stellar-mass black hole population (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our Gaussian
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

@@

Small values of $\chieff$ as exhibited in these systems can result
from either intrinsically small spins or larger spins whose direction
is mis-aligned with the orbital plane of the binary (i.e.\ spin
vectors with small $z$-components).  Mis-alignment is capable of
producing \emph{negative} values of $\chieff$, however, whereas aligned
spins will always have $\chieff \geq 0$.  This difference provides
strong discriminating power between the two angular distributions,
even without good information about the magnitude distribution; to the
extent that data favour negative $\chieff$ they weigh heavily against
aligned models.  To quantify the degree of support for these two
alternate explanations of small $\chieff$ values in the merging
\ac{BBH} population, we compared the Bayesian evidence for various
simple models of the spin population using the \ac{GW} data set.

@@

\begin{figure}
  \plotone{../plots/chi-eff-distributions}
  \caption{\label{fig:chieff-distribution-models} The models for the
    distribution of $\chieff$ considered in this paper.  In all models
    we assume that the binary mass ratio $q \equiv m_1/m_2 = 1$ and
    that the distribution of spin vectors is the same for each
    component.  The ``flat'' (blue lines), ``high,'' (green lines),
    and ``low'' (red lines) magnitude distributions are defined in
    Eq.\ \eqref{eq:magnitude-dists}.  Solid lines give the $\chieff$
    distribution under the assumption that the orientations of the
    spins are isotropic; dashed lines give the distribution under the
    assumption that both objects' spins are aligned with the orbital
    angular momentum.  The isotropic distributions are readily
    distinguished from the aligned distributions by the production of negative
    $\chieff$ values, while the distinction between the three models
    for the spin magnitude distribution is less sharp.}
\end{figure}

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming an
additional 10 detections with similar observational uncertainties
drawn from each of our six zero-parameter models for the spin
distribution in addition to the three existing detections from O1, we
find the odds ratios shown in Figure \ref{fig:O2-predictions}.  We
find that most scenarios with an additional 10 detections allow the
correct angular distribution to be distinguished with greater than
$5\sigma$ ($2.9 \times 10^{-7}$ odds) credibility; and in the most
pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we will remain much less certain
about the \emph{magnitude} distribution until we have a larger number
of observations.  In Figure \ref{fig:O2-predictions}, the odds ratio
between different magnitude distributions with the same angular
distribution is much closer to unity than the odds ratio between angular
distributions.

@@

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments
\citep{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016}
are expected to be isotropic given the absence of a preferred
direction \citep[e.g.,][]{2016ApJ...832L...2R} and the persistence of
an isotropic distribution through post-Newtonian evolution
\citep{2004PhRvD..70l4020S,2007ApJ...661L.147B}.

The spin directions in isolated binaries, whether evolving via the
classical channel through a common-envelope phase
\citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017}
or through chemically homogeneous evolution
\citep{MandeldeMink:2016,Marchant:2016} are usually expected to be
preferentially aligned.  Despite observed spin-orbit misalignments in
both massive stellar binaries
\citep{Albrecht:2009,2014ApJ...785...83A} and BH XRBs
\citep{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014},
mass accretion and tidal interactions will tend to realign the binary.
On the other hand, a supernova natal kick (if any) can change the
orbital plane and misalign the binary
\citep{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can
also tilt the spin angle, as in the double pulsar J0737-3039
\citep{2011ApJ...742...81F}; and a variety of uncertain processes,
such as wind-driven mass loss and post-collapse fallback, can couple
the spin magnitude and direction distributions, contrary to our
simplified assumptions.
